---
title: "labor_updated"
author: "Allie Bauman"
date: '2022-08-19'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this file we get all of the data associated with the labor 2-pager. All data is from the U.S. Bureau of Labor Statistics Quarterly Census of Employment and Wages, 2021. 

https://www.bls.gov/cew/downloadable-data-files.htm

NAICS codes of interest include: 
  * Crop Production 111
  * Animal Production 112
  * Fishing, Hunting, and Trapping 114 
  * Support Activities for Crop and Animal Production 115
  * Food manufacturing 311 
  * Animal Food Processing 3111
  * Grain and Oilseed Milling 3112
  * Sugar and Confectionery Product Manufacturing 3113
  * Fruit and Vegetable Preserving and Specialty Food Manufacturing 3114
  * Dairy Product Manufacturing 3115
  * Animal Slaughtering and Processing 3116
  * Seafood Product Preparation and Packaging 3117 
  * Bakeries and Tortilla Manufacturing (except Retail Bakeries) 3118
  * Other Food Manufacturing 3119
  * Beverage Manufacturing 3121
  * Tobacco Manufacturing 3122
  * Pesticide and Other Agricultural Chemical Manufacturing 32532 
  * Agricultural Implement Manufacturing 33311 
  * Grocery and Related Product Merchant Wholesalers 4244
  * Farm Product Raw Materials, Merchant Wholesalers 4245
  * Beer, Wine, and Distilled Alcoholic Beverage Merchant Wholesalers 4248 
  * Farm Supplies, Merchant Wholesalers 42491
  * Truck Transportation 484
  * Warehousing and Storage 493 
  * Food and Beverage Stores 445
  * Grocery Stores 4451 
  * Specialty Food Stores 4452
  * Beer, Wine, and Liquor Stores 4453
  * Supermarkets and other grocery stores 44511 
  * Convenience stores 44512 
  * Retail Bakeries 311811 
  * Food Service and Drinking Places 722
  * Special Food Services 7223
  * Drinking Places (Alcoholic Beverages) 7224
  * Restaurants and Other Eating Places 7225
  * Full service restaurants 722511 
  * Caterers 722320 
  * Mobile food services 722330 
  

I put all relevant NAICS codes into one folder and import all contents of the folder and bind into one data frame. 

We compute data for the following

  * privately held businesses (own_title == "Private")
  * all establishment sizes (size_code==0)
  * annual data (qtr=="A")
  * remove data with disclosure, otherwise they are included in the data as a zero, we want NA
  * remove area fips starting with "C" as these are for MSA's

```{r}
library(tidyverse)
library(janitor)

# Import and bind all NAICS codes of interest
data_path <- "data/QCEW/food"

files <- dir(data_path, pattern = "*.csv")
labor <- files %>% 
  map(~read_csv(file.path(data_path, .))) %>%
  reduce(bind_rows) %>% clean_names() 

# Separate fips so we can identify state level data
labor <- labor %>% separate(area_fips, 
                             c("state_fips", "county_fips"),
                             sep = 2,
                             remove = FALSE)

# Select data of interest
labor <- labor %>% filter(own_title=="Private" & size_code==0 & is.na(disclosure_code) &
                            qtr=="A" & !str_detect(area_fips, "^C")) %>% select(
                              area_fips, state_fips, county_fips, year, 
                              industry_code, industry_title, 
                              total_annual_wages, annual_avg_emplvl, 
                              annual_avg_estabs_count, annual_avg_wkly_wage, lq_annual_avg_emplvl)

rm(data_path, files)
```

## Wages (in thousands) for relevant NAICS codes 

```{r}

# Select wage data 
wages <- labor %>% select(area_fips, state_fips, county_fips, year, 
                          industry_code, industry_title, total_annual_wages)
  
# Combine NAICS and variable name, add category and topic area
wages <- wages %>% mutate(
  variable_name = "total_annual_wages", 
  category = "Labor", 
  topic_area = "Wages") %>% unite(
    "variable_name", c(variable_name, industry_title)) %>% rename(
      value = total_annual_wages)

# Change state level FIPS codes to be 2 digits so are consistent with the rest of the data 
wages <- wages %>% mutate(
  fips = case_when(
    county_fips =="000" & state_fips!="US" ~ state_fips, 
    county_fips !="000"  ~ area_fips,
    state_fips=="US" ~ "00")) %>% select(
      fips, topic_area, category, year, variable_name, value)

```

## Annual average of monthly employment levels for a given year

```{r}

# Select employment data
employment <- labor %>% select(area_fips, state_fips, county_fips, year, 
                          industry_code, industry_title, annual_avg_emplvl)
  
# Combine NAICS and variable name, add category and topic area
employment <- employment %>% mutate(
  variable_name = "annual_avg_emplvl", 
  category = "Labor", 
  topic_area = "Employment") %>% unite(
    "variable_name", c(variable_name, industry_title)) %>% rename(
      value = annual_avg_emplvl)

# Change state level FIPS codes to be 2 digits so are consistent with the rest of the data 
employment <- employment %>% mutate(
  fips = case_when(
    county_fips =="000" & state_fips!="US" ~ state_fips, 
    county_fips !="000"  ~ area_fips,
    state_fips=="US" ~ "00")) %>% select(
      fips, topic_area, category, year, variable_name, value)

```

## Annual average of quarterly establishment counts for a given year

```{r}

# Select employment data
establishments <- labor %>% select(area_fips, state_fips, county_fips, year, 
                          industry_code, industry_title, annual_avg_estabs_count)
  
# Combine NAICS and variable name, add category and topic area
establishments <- establishments %>% mutate(
  variable_name = "annual_avg_estabs_count", 
  category = "Labor", 
  topic_area = "Employment") %>% unite(
    "variable_name", c(variable_name, industry_title)) %>% rename(
      value = annual_avg_estabs_count)

# Change state level FIPS codes to be 2 digits so are consistent with the rest of the data 
establishments <- establishments %>% mutate(
  fips = case_when(
    county_fips =="000" & state_fips!="US" ~ state_fips, 
    county_fips !="000"  ~ area_fips,
    state_fips=="US" ~ "00")) %>% select(
      fips, topic_area, category, year, variable_name, value)

```

## Average weekly wage based on the 12-monthly employment levels and total annual wage levels.

```{r}

# Select employment data
weekly_wage <- labor %>% select(area_fips, state_fips, county_fips, year, 
                          industry_code, industry_title, annual_avg_wkly_wage)
  
# Combine NAICS and variable name, add category and topic area
weekly_wage <- weekly_wage %>% mutate(
  variable_name = "annual_avg_wkly_wage", 
  category = "Labor", 
  topic_area = "Wages") %>% unite(
    "variable_name", c(variable_name, industry_title)) %>% rename(
      value = annual_avg_wkly_wage)

# Change state level FIPS codes to be 2 digits so are consistent with the rest of the data 
weekly_wage <- weekly_wage %>% mutate(
  fips = case_when(
    county_fips =="000" & state_fips!="US" ~ state_fips, 
    county_fips !="000"  ~ area_fips,
    state_fips=="US" ~ "00")) %>% select(
      fips, topic_area, category, year, variable_name, value)

```

## Location quotient

There are different ways to define location quotient. We use location quotient of annual average employment relative to the U.S. (Rounded to the hundredths place). This is the same metric as used by IMPLAN. 

```{r}

# Select lq data
lq <- labor %>% select(area_fips, state_fips, county_fips, year, 
                          industry_code, industry_title, lq_annual_avg_emplvl)
  
# Combine NAICS and variable name, add category and topic area
lq <- lq %>% mutate(
  variable_name = "lq_annual_avg_emplvl", 
  category = "Labor", 
  topic_area = "Employment") %>% unite(
    "variable_name", c(variable_name, industry_title)) %>% rename(
      value = lq_annual_avg_emplvl)

# Change state level FIPS codes to be 2 digits so are consistent with the rest of the data 
lq <- lq %>% mutate(
  fips = case_when(
    county_fips =="000" & state_fips!="US" ~ state_fips, 
    county_fips !="000"  ~ area_fips,
    state_fips=="US" ~ "00")) %>% select(
      fips, topic_area, category, year, variable_name, value)

```

## Combine all data into one data frame 

We remove data without a county name. The QCEW data has some fips codes called Oversees locations, Multicounty, out-of-state, and unknown or undefined. These fips codes do not have a match in the more county-level fips codes, so we drop all observations without a county name in order to drop these observations. 

```{r}
library(tidycensus)

# Get county and state fips, state name, county name 
county <- tidycensus::fips_codes %>% unite("fips", c(state_code, county_code), sep = "", remove = FALSE) %>% rename(county_name = county) %>% select(fips, county_name, state_name)

state <- tidycensus::fips_codes %>% select(state_code, state_name) %>% 
  rename(fips = state_code, 
         state_name1 = state_name) %>% distinct()

# Bind wages, employment, establishments, weekly_wage data
labor <- bind_rows(wages, employment, establishments, weekly_wage, lq) %>% 
  left_join(county) %>% left_join(state)

# Coalesce state and county names so state name is present in all rows
labor <- labor %>% mutate(
    state_name = coalesce(labor$state_name, labor$state_name1)) %>% select(
      fips, county_name, state_name, category, topic_area, year, variable_name, value) %>% 
  filter(!(is.na(county_name) & is.na(state_name)))

# make year a character vector to be consistent with other data sets
labor$year <- as.character(labor$year)

rm(wages, employment, establishments, weekly_wage, lq)
```

# Sex by occupation 

Data from U.S.Census Bureau American Community Survey Table, B24010, SEX BY OCCUPATION FOR THE CIVILIAN EMPLOYED POPULATION 16 YEARS AND OVER, 2019 1-year estimate.

https://data.census.gov/cedsci/table?q=B24010

Variables include 

  * B24010_011E	Estimate!!Total:!!Male:!!Management, business, science, and arts occupations:!!Management, business, and financial occupations:!!Management occupations:!!Farmers, ranchers, and other agricultural managers
  * B24010_076E	Estimate!!Total:!!Male:!!Service occupations:!!Food preparation and serving related occupations:
  * B24010_113E	Estimate!!Total:!!Male:!!Natural resources, construction, and maintenance occupations:!!Farming, fishing, and forestry occupations:
  * B24010_162E	Estimate!!Total:!!Female:!!Management, business, science, and arts occupations:!!Management, business, and financial occupations:!!Management occupations:!!Farmers, ranchers, and other agricultural managers
  * B24010_227E	Estimate!!Total:!!Female:!!Service occupations:!!Food preparation and serving related occupations:
  * B24010_264E	Estimate!!Total:!!Female:!!Natural resources, construction, and maintenance occupations:!!Farming, fishing, and forestry occupations:
  
```{r}

# Import metadata
var_defs_gender <- read_csv("data/ACS/ACSDT1Y2019.B24010-Column-Metadata.csv") %>%
  clean_names() %>% filter(
  str_detect(label, "^Estimate!!") & 
    (str_detect(label, "!!Farmers, ranchers, and other agricultural managers$") | 
    str_detect(label, "!!Food preparation and serving related occupations:$") | 
    str_detect(label, "!!Farming, fishing, and forestry occupations:$")))

write_csv(var_defs_gender, "data/ACS/var_defs_gender.csv")
rm(var_defs_gender)


# Import and bind all NAICS codes of interest
data_path <- "data/ACS/gender_data"

files <- dir(data_path, pattern = "*.csv")
gender <- files %>% 
  map(~read_csv(file.path(data_path, .)) %>% slice(-1)) %>%
  reduce(bind_rows) 

# Import data and keep columns of interest, get fips from GEO_ID
gender <- gender %>% select(
  GEO_ID, NAME,
  B24010_011E, B24010_076E, B24010_113E, 
  B24010_162E, B24010_227E, B24010_264E) %>% mutate(
    fips = case_when(
      	str_detect(GEO_ID, "0400000") ~ str_sub(GEO_ID,-2), 
        str_detect(GEO_ID, "0100000") ~ "US", 
        str_detect(GEO_ID, "0500000") ~ str_sub(GEO_ID, -5))) 

# Rename data and add columns to match other data
gender <- gender %>%  rename(
  number_male_farmers_ranchers_ag_managers = B24010_011E, 
  number_male_food_preparation_serving_occupations = B24010_076E, 
  number_male_farming_fishing_forestry_occupations = B24010_113E, 
  number_female_farmers_ranchers_ag_managers = B24010_162E, 
  number_female_food_preparation_serving_occupations = B24010_227E, 
  number_female_farming_fishing_forestry_occupations = B24010_264E) %>% select(-GEO_ID)

# Pivot longer
gender <- gender %>% pivot_longer(
  cols = -c(fips, NAME), 
  names_to = "variable_name", 
  values_to = "value")
  
# Add in state/county and variables and put in correct order
gender <- gender %>% left_join(county) %>% mutate(
  category = "Labor", 
  topic_area = "Employment", 
  year = "2019") %>% mutate(
    state_name = coalesce(NAME, state_name), 
    state_name = ifelse(state_name=="United States", NA, state_name)) %>%
  
  select(
    fips, county_name, state_name, category, topic_area, year, variable_name, value)

# Replace "NULL" with missing and drop missing rows and make numeric
gender <- gender %>% mutate(
  value = ifelse(value=="null", NA, value)) %>% filter(
    !is.na(value))

gender$value <- as.numeric(gender$value)

```

# Median earnings by sector and gender 

We use 2020 5-Year Estimates data from the U.S. Census Bureau, American Community Survey, S2411  OCCUPATION BY SEX AND MEDIAN EARNINGS IN THE PAST 12 MONTHS (IN 2020 INFLATION-ADJUSTED DOLLARS) FOR THE CIVILIAN EMPLOYED POPULATION 16 YEARS AND OVER

https://data.census.gov/cedsci/table?q=S2411%3A%20OCCUPATION%20BY%20SEX%20AND%20MEDIAN%20EARNINGS%20IN%20THE%20PAST%2012%20MONTHS%20%28IN%202019%20INFLATION-ADJUSTED%20DOLLARS%29%20FOR%20THE%20CIVILIAN%20EMPLOYED%20POPULATION%2016%20YEARS%20AND%20OVER&g=0100000US,%240500000&y=2020&tid=ACSST5Y2020.S2411

Data notes: there are some observations with a + at the end or - at the beginning, making them non-numeric. This will not work with the rest of the data. We remove the +/- sign from these numbers and convert the value column to numeric. 

We include two industries: 

* Food preparation and serving related occupations 
* Farming, fishing, and forestry occupations 

We include median earning for men, women and women's earnings as a percent of men's earnings. 

Variables include: 

  * S2411_C02_023E	Estimate!!Median earnings (dollars) for male!!Civilian employed population 16 years and over with earnings!!Service occupations:!!Food preparation and serving related occupations
  * S2411_C02_030E	Estimate!!Median earnings (dollars) for male!!Civilian employed population 16 years and over with earnings!!Natural resources, construction, and maintenance occupations:!!Farming, fishing, and forestry occupations
  * S2411_C03_023E	Estimate!!Median earnings (dollars) for female!!Civilian employed population 16 years and over with earnings!!Service occupations:!!Food preparation and serving related occupations
  * S2411_C03_030E	Estimate!!Median earnings (dollars) for female!!Civilian employed population 16 years and over with earnings!!Natural resources, construction, and maintenance occupations:!!Farming, fishing, and forestry occupations
  * S2411_C04_023E	Estimate!!Women's earnings as a percentage of men's earning!!Civilian employed population 16 years and over with earnings!!Service occupations:!!Food preparation and serving related occupations
  * S2411_C04_030E	Estimate!!Women's earnings as a percentage of men's earning!!Civilian employed population 16 years and over with earnings!!Natural resources, construction, and maintenance occupations:!!Farming, fishing, and forestry occupations

```{r}

# Import metadata - only have columns of interest
var_defs_earnings <- read_csv("data/ACS/ACSST5Y2020.S2411-Column-Metadata.csv") %>% 
  clean_names() %>% filter(
  str_detect(label, "^Estimate!!") &    
  (str_detect(label, "!!Food preparation and serving related occupations$") | 
     str_detect(label, "!!Farming, fishing, and forestry occupations$")))

write_csv(var_defs_earnings, "data/ACS/var_defs_earnings.csv")
rm(var_defs_earnings)

# Import and bind all NAICS codes of interest
data_path <- "data/ACS/median_income_data"

files <- dir(data_path, pattern = "*.csv")
earnings <- files %>% 
  map(~read_csv(file.path(data_path, .)) %>% slice(-1)) %>%
  reduce(bind_rows) 

# Import data and keep columns of interest, get fips from GEO_ID
earnings <- earnings %>% select(
  GEO_ID, NAME,
  S2411_C02_023E, S2411_C02_030E, S2411_C03_023E, 
  S2411_C03_030E, S2411_C04_023E, S2411_C04_030E) %>% mutate(
    fips = case_when(
      	str_detect(GEO_ID, "0400000") ~ str_sub(GEO_ID,-2), 
        str_detect(GEO_ID, "0100000") ~ "US", 
        str_detect(GEO_ID, "0500000") ~ str_sub(GEO_ID, -5))) 

# Rename data and add columns to match other data
earnings <- earnings %>%  rename(
  median_earnings_male_food_preparation_serving = S2411_C02_023E, 
  median_earnings_male_farming_fishing =  S2411_C02_030E, 
  median_earnings_female_food_preparation_serving = S2411_C03_023E, 
  median_earnings_female_farming_fishing = S2411_C03_030E, 
  median_earnings_female_percent_male_food_preparation_serving = S2411_C04_023E, 
  median_earnings_female_percent_male_farming_fishing = S2411_C04_030E) %>% select(-GEO_ID)

# Pivot longer
earnings <- earnings %>% pivot_longer(
  cols = -c(fips, NAME), 
  names_to = "variable_name", 
  values_to = "value")
  
# Add in state/county and variables and put in correct order
earnings <- earnings %>% left_join(county) %>% mutate(
  category = "Labor", 
  topic_area = "Wages", 
  year = "2020") %>% mutate(
    state_name = coalesce(NAME, state_name), 
    state_name = ifelse(state_name=="United States", NA, state_name)) %>% select(
    fips, county_name, state_name, category, topic_area, year, variable_name, value)

# Replace "-" with missing and drop missing rows
earnings <- earnings %>% mutate(
  value = ifelse(value=="-", NA, value)) %>% filter(
    !is.na(value))

# Remove "+" and "," from 250,000+ and convert value to numeric
earnings <- earnings %>% mutate(
  value = ifelse(value=="250,000+", 250000, value), 
  value = ifelse(value=="2,500-", 2500, value))

earnings$value <- as.numeric(earnings$value)

```

## Farm labor
Download U.S. county-level data from the 2017 Census of Agriculture using the Census Data Query Tool. 

https://www.nass.usda.gov/Publications/AgCensus/2017/index.php

Note that this data set it too large to store on GitHub, so the only files that are stored are those relevant to this analysis. Farm labor data is available in Census Volume 1, Chapter 2: County level data, Table 7: Hired farm labor. 

County - fips 02010 ALEUTIAN ISLANDS is not a fips code found in the other data sets. 

```{r}

# Import data and keep data of interest only
# Only run this if you download the full data file from the Census website
#farm_labor <- coa_17 <- read_tsv("data/2017_cdqt_data.txt") %>% clean_names() %>% filter(
#  census_chapter==2 & census_table==7 & 
#    (census_row==2 | census_row == 51 | census_row ==55)) 

#write_csv(farm_labor, "data/farm_labor.csv")

# Manipulate data so we can add state/county names
read_csv("data/farm_labor.csv")
farm_labor <- farm_labor %>% select(
  state_fips_code, county_code, state_name, short_desc, value) %>% unite(
    "fips", state_fips_code:county_code, sep = "", remove = FALSE) %>% mutate(
      fips = ifelse(county_code=="NULL", NA, fips)) %>% mutate(
      fips = coalesce(fips,state_fips_code), 
      fips = ifelse(fips=="99", "US", fips)) 

# Import county data, remove state name
county <- tidycensus::fips_codes %>% unite("fips", c(state_code, county_code), sep = "", remove = FALSE) %>% rename(county_name = county) %>% select(fips, county_name)

# Use state name from NASS data, but make lower case to match other state name data
farm_labor <- farm_labor %>% left_join(county, by = "fips") %>% mutate(
  state_name = tools::toTitleCase(tolower(state_name)),
  state_name = ifelse(fips=="US", NA, state_name)) 

# Change national fips code to be "US" to match with other data sets, rename variables, add new columns and keep only what we need
farm_labor <- farm_labor %>% mutate(
  variable_name = case_when(
    short_desc=="LABOR, HIRED - NUMBER OF WORKERS" ~ "hired_labor_number_workers", 
    short_desc=="LABOR, MIGRANT - NUMBER OF WORKERS" ~ "migrant_labor_number_workers", 
    short_desc=="LABOR, UNPAID - NUMBER OF WORKERS" ~ "unpaid_labor_number_workers"), 
  category = "Labor", 
  topic_area = "Employment", 
  year = "2017") %>% select(
    fips, county_name, state_name, category, topic_area, year, variable_name, value)
    
```

## Join all labor data and write metadata file

```{r}

# Join data and write data file
labor <- bind_rows(labor, gender, earnings, farm_labor)

write_csv(labor, "labor.csv")

# Write metadata file
metadata <- labor %>% distinct(variable_name, .keep_all = TRUE) %>% select(category, topic_area, variable_name)

# Add columns
metadata <- metadata %>% mutate(
  `2 pager title` = "Labor", 
  user_friendly_variable_name = case_when(
    str_detect(variable_name, "^total_annual_wages") | 
      str_detect(variable_name, "^median_earnings") | 
      str_detect(variable_name, "^number_") ~ variable_name,
    str_detect(variable_name, "^annual_avg_emplvl") ~ 
      str_replace_all(variable_name, 
                      pattern = "^annual_avg_emplvl", 
                      replacement = "annual_average_employment_level"),
    str_detect(variable_name, "^annual_avg_estabs_count") ~ 
      str_replace_all(variable_name, 
                      pattern = "^annual_avg_estabs_count", 
                      replacement = "number_of_establishments"), 
    str_detect(variable_name, "^annual_avg_wkly_wage") ~ 
      str_replace_all(variable_name, 
                      pattern = "^annual_avg_wkly_wage", 
                      replacement = "average_weekly_wage"), 
    str_detect(variable_name, "^lq_annual_avg_emplvl") ~
      str_replace_all(variable_name, 
                      pattern = "^lq_annual_avg_emplvl", 
                      replacement = "location_quotient"),
    str_detect(variable_name, "number_workers$") ~ 
      str_replace_all(variable_name, 
                      pattern = "number_workers$", 
                      replacement = "number_of_workers"))) %>% 
  select(`2 pager title`, everything())

write_csv(metadata, "metadata_labor1.csv")
```