---
title: "Local Food Sales"
author: "Allie Bauman"
date: '2022-08-25'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this file, we import and manipulate all data related to Local Food Sales.  


## Community supported agriculture businesses and Farmers Market

We gather point level data from the USDA Local Food Directories, CSA Enterprise Directory and Farmers Market Directory. We match the lat/long coordinates to county level data to provide counts of CSAs by county, state, and US. 

https://www.usdalocalfoodportal.com/#directories

https://www.ams.usda.gov/local-food-directories/csas
https://www.usdalocalfoodportal.com/fe/fdirectory_farmersmarket/?source=fe&directory=farmersmarket&location=&x=&y=


```{r}
library(tidyverse)
library(sf)

# Import data, keep columns of interest and drop obs. with NA for lat/long
csa <- readxl::read_xlsx("data/csa_2022-825163837.xlsx") %>% select(
  listing_id, listing_name, location_address, location_x, location_y) %>% mutate(
    variable_name = "number_CSAs")

fm <- readxl::read_xlsx("data/farmersmarket_2022-825165232.xlsx") %>% select(
  listing_id, listing_name, location_address, location_x, location_y) %>% mutate(
    variable_name = "number_farmers_markets")

fm_csa <- bind_rows(csa, fm) %>% filter(!is.na(location_x), !is.na(location_y)) %>% 
  rename(lat = location_y, long = location_x)
rm(csa, fm)

## Make point level data set ----STILL NEED TO DO 

## Get FIPS codes from lat/long data
# Get county spatial data frame
counties <- USAboundaries::us_counties(resolution = "high")

# make data frame into a spatial data frame
fm_csa_agg <- fm_csa %>% st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

# Join point level meat data to county data, return the fips code, and turn back into a regular data frame
fm_csa_agg <- st_join(fm_csa_agg, left = FALSE, counties["geoid"]) %>% as.data.frame() %>% 
  rename(fips = geoid)

# Number per county - Add back in missing counties so we can have NA values for them
state <- tidycensus::fips_codes %>% unite("fips", c(state_code, county_code), sep = "", remove = FALSE) %>% select(fips, state_code)

count <- fm_csa_agg %>% group_by(fips, variable_name) %>% count() %>% rename(value = n)
count <- full_join(state, count) %>% select(!state_code) %>% filter(!is.na(value))

fm_csa_county <- count %>% mutate(
  category = "Food Retail",
  topic_area = case_when(
    variable_name == "number_CSAs" ~ "Community Supported Agriculture", 
    variable_name == "number_farmers_markets" ~ "Farmers Market"),
  year = "2022") %>% select(
    fips, topic_area, category, year, variable_name, value) 

# Number per state
count <- left_join(fm_csa_agg, state) %>% group_by(state_code, variable_name) %>% 
  count() %>% rename(
  value = n, 
  fips = state_code)

fm_csa_state <- count %>% mutate(
  category = "Food Retail",
  topic_area = case_when(
    variable_name == "number_CSAs" ~ "Community Supported Agriculture", 
    variable_name == "number_farmers_markets" ~ "Farmers Market"),
  year = "2022") %>% select(
    fips, topic_area, category, year, variable_name, value) 

# Number US
count <- fm_csa_agg %>% group_by(variable_name) %>% count() %>% rename(value = n)

fm_csa_us<- count %>% mutate(
  fips = "US",
  category = "Food Retail",
  topic_area = case_when(
    variable_name == "number_CSAs" ~ "Community Supported Agriculture", 
    variable_name == "number_farmers_markets" ~ "Farmers Market"),
  year = "2022") %>% select(
    fips, topic_area, category, year, variable_name, value) 

# bind into one data frame 
fm_csa_agg <- bind_rows(fm_csa_county, fm_csa_state, fm_csa_us)
rm(fm_csa_county, fm_csa_state, fm_csa_us, count)

rm(counties, state)
```

## Combine all data and put in final format 

```{r}
fips <- tidycensus::fips_codes %>% unite("fips", c(state_code, county_code), sep = "", remove = FALSE) %>% select(fips, state_name, county) %>% rename(county_name= county) 

local <- fm_csa_agg %>% left_join(fips) %>% select(
  fips, county_name, state_name, category, topic_area, year, variable_name, value)

# Write file 
write_csv(local, "local_food_sales.csv")
```
