---
title: "Local Food Sales"
author: "Allie Bauman"
date: '2022-08-25'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this file, we import and manipulate all data related to Local Food Sales.  


## Community supported agriculture businesses and Farmers Market

We gather point level data from the USDA Local Food Directories, CSA Enterprise Directory and Farmers Market Directory. We match the lat/long coordinates to county level data to provide counts of CSAs by county, state, and US. 

https://www.usdalocalfoodportal.com/#directories

https://www.ams.usda.gov/local-food-directories/csas
https://www.usdalocalfoodportal.com/fe/fdirectory_farmersmarket/?source=fe&directory=farmersmarket&location=&x=&y=

We only keep data that was updated since 8/30/2020 (based on feedback from AMS). If a county has no reported farmers markets or CSAs in this database, our final data will show there are zero in the county. 

```{r}
library(tidyverse)
library(sf)
library(lubridate)

# Import data, keep columns of interest, drop obs. with NA for lat/long, drop obs. not updated in last two years
csa <- readxl::read_xlsx("data/csa_2022-825163837.xlsx") %>%
  mutate(
    variable_name = "number_CSAs", 
    update_time = as_date(update_time)) %>% filter(
    update_time>"2020-08-30", 
    !is.na(location_x), !is.na(location_y)) %>% 
  select(
    listing_id, variable_name, listing_name, location_address, 
    location_x, location_y)

fm <- readxl::read_xlsx("data/farmersmarket_2022-825165232.xlsx") %>%
  mutate(
    variable_name = "number_farmers_markets", 
    update_time = as_date(update_time)) %>% filter(
    update_time>"2020-08-30", 
    !is.na(location_x), !is.na(location_y)) %>% 
  select(
    listing_id, variable_name, listing_name, location_address, 
    location_x, location_y)

fm_csa <- bind_rows(csa, fm) %>% 
  rename(lat = location_y, long = location_x)
rm(csa, fm)

## Get FIPS codes from lat/long data
# Get county spatial data frame
counties <- USAboundaries::us_counties(resolution = "high")

# make data frame into a spatial data frame
fm_csa_agg <- fm_csa %>% st_as_sf(coords = c("long", "lat"), crs = 4326, remove = FALSE)

# Join point level meat data to county data, return the fips code, and turn back into a regular data frame
fm_csa_agg <- st_join(fm_csa_agg, left = FALSE, counties["geoid"]) %>% as.data.frame() %>% 
  rename(fips = geoid)

# Number per county - Add back in missing counties so we can have a 0 value for those counties

## Get state level fips codes
state <- tidycensus::fips_codes %>% unite("fips", c(state_code, county_code), sep = "", remove = FALSE) %>% select(fips, state_code)

## Count number per county
count <- fm_csa_agg %>% group_by(fips, variable_name) %>% count() %>% rename(value = n)

## We need to have a zero for all counties not listed in the data (not an NA). Create a data frame with two rows per county, one with variable names number_farmers_markets, and the other number_CSAs. 
state_fm <- state %>% mutate(
  variable_name = "number_farmers_markets") 

state_csa <- state %>% mutate(
  variable_name = "number_CSAs")

state_updated <- bind_rows(state_fm, state_csa) %>% arrange(fips)

## Join number per county with county fips data, replace NAs with zero
count <- full_join(state_updated, count) %>% 
  select(!state_code) %>% mutate(
    value = ifelse(is.na(value), 0, value))

fm_csa_county <- count %>% mutate(
  category = "Food Retail",
  topic_area = case_when(
    variable_name == "number_CSAs" ~ "Community Supported Agriculture", 
    variable_name == "number_farmers_markets" ~ "Farmers Market"),
  year = "2022") %>% select(
    fips, topic_area, category, year, variable_name, value) 

# Number per state
count <- left_join(fm_csa_agg, state) %>% group_by(state_code, variable_name) %>% 
  count() %>% rename(
  value = n, 
  fips = state_code)

fm_csa_state <- count %>% mutate(
  category = "Food Retail",
  topic_area = case_when(
    variable_name == "number_CSAs" ~ "Community Supported Agriculture", 
    variable_name == "number_farmers_markets" ~ "Farmers Market"),
  year = "2022") %>% select(
    fips, topic_area, category, year, variable_name, value) 

# Number US
count <- fm_csa_agg %>% group_by(variable_name) %>% count() %>% rename(value = n)

fm_csa_us<- count %>% mutate(
  fips = "00",
  category = "Food Retail",
  topic_area = case_when(
    variable_name == "number_CSAs" ~ "Community Supported Agriculture", 
    variable_name == "number_farmers_markets" ~ "Farmers Market"),
  year = "2022") %>% select(
    fips, topic_area, category, year, variable_name, value) 

# bind into one data frame 
fm_csa_agg <- bind_rows(fm_csa_county, fm_csa_state, fm_csa_us)
rm(fm_csa_county, fm_csa_state, fm_csa_us, count, fm_csa)

rm(counties)
```

## Import COA data
Here we import data compiled from the restricted access 2017 Census of Agriculture data on the number of farms participating in local food marketing channels by county. 

There is one county in the data set that does not have a match in the rest of the data (fips 02010). We drop this observation. 

```{r}

# Import data
coa <- readxl::read_xlsx(
  "data/08.08.22_AMS Result tables_suppressed.xlsx", 
  sheet = "market_channel") %>% mutate(
    fips = str_pad(fips, side = "left", pad = "0",
                   width = 5)) %>% filter(!fips=="02010")

# calculate percentage variables 
coa <- coa %>% mutate(across(d2c_only:nonlocal, 
                             .fns = ~./n, 
                             .names = "{col}_p"))

# make data long and add year variable
coa_county <- coa %>% rename(all_farms = n) %>% pivot_longer(
  cols = !fips, 
  names_to = "variable_name", 
  values_to = "value") %>% mutate(
    year = "2017")

# Aggregate to the state level
coa_state <- coa %>% left_join(state) %>% group_by(state_code) %>% 
  summarise(across(n:nonlocal, ~ sum(.))) %>% 
  mutate(across(d2c_only:nonlocal, 
                             .fns = ~./n, 
                             .names = "{col}_p"))

coa_state <- coa_state %>% rename(
  all_farms = n, 
  fips = state_code) %>% pivot_longer(
  cols = !fips, 
  names_to = "variable_name", 
  values_to = "value") %>% mutate(
    year = "2017") 

# Aggregate to the national level 
coa_US <- coa %>%  
  summarise(across(n:nonlocal, ~ sum(.))) %>% 
  mutate(across(d2c_only:nonlocal, 
                             .fns = ~./n, 
                             .names = "{col}_p"))

coa_US <- coa_US %>% rename(
  all_farms = n) %>% mutate(
    year = "2017", 
    fips = "00") %>% pivot_longer(
      cols = !c(year, fips), 
      names_to = "variable_name", 
      values_to = "value") 

# Bind rows 
coa <- bind_rows(coa_county, coa_state, coa_US)

# Add columns 
coa <- coa %>% mutate(
  category = "Food Retail",
  topic_area = "General DTC") %>% select(
    fips, topic_area, category, year, variable_name, value) 

rm(coa_county, coa_state, coa_US, state)

# write file for metadata 
coa_meta <- coa %>% select(category, topic_area, variable_name) %>% distinct()
write_csv(coa_meta, "coa_meta_temp.csv")

```

## Combine all data and put in final format 

```{r}
fips <- tidycensus::fips_codes %>% unite("fips", c(state_code, county_code), sep = "", remove = FALSE) %>% select(fips, state_name, county) %>% rename(county_name= county) 

local <- bind_rows(fm_csa_agg, coa) %>% left_join(fips) %>% select(
  fips, county_name, state_name, category, topic_area, year, variable_name, value)

# Write file 
write_csv(local, "local_food_sales.csv")
```
