---
title: "Capitals - updated"
author: "Allie Bauman"
date: '2022-07-13'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Indicators of Community Wealth 

We first import the data set and modify the data as described in Schmitt et al. (2021) do file to make variables per capita, etc. We keep the variables of interest, add a year variable and organize data by capital type into separate csv files. 

We use the CapitalPCs file from the GitHub site for the Schmitt et al. (2021) paper that developed these indicators to obtain the principal components. https://github.com/schmi-ny/County-Level-Community-Capital-Stocks/blob/main/CapitalPCs061521.xlsx

These data and capitals were then updated to include additional years of data. While we do include the additional years of data here, the principal component measures of capitals are based on the original data, as provided in Schmitt et al. (2021). 

The file called "capitals_new.csv" contains all of the oringial data used in Schmitt et al. (2021) plus additional years of older data that was collected based on data availability. We use the data in this file and join the principal component measures of capitals from Schmitt et al. (2021) to provide the final data set for each capital type.

Where possible, we will aggregate data by the state and national level to be used for comparison. All aggregation will take place before variables are converted to per capita or similar measures. Principal components will not be available at the state or national level. 

```{r}
library(tidyverse)
library(readxl)
library(tidycensus)

# Import data
aeppPCs <- read_xlsx("CapitalPCs061521.xlsx") %>% select(
  fips1, pc1b_manufacturing:pc2s_publicvoiceparticipation) %>% mutate(
  fips1 = str_pad(fips1, pad = "0", width = 5, side = "left"))

capitals <- read_csv("capitals_new.csv") %>% mutate(
  fips1 = str_pad(fips1, pad = "0", width = 5, side = "left"))

# Join principal components to capitals data
capitals <- left_join(capitals, aeppPCs, by = "fips1") 

# Keep fips1 and rename as fips2
capitals <- capitals %>% select(-c(fips, fips2)) %>% rename(
    fips = fips1) 

# Get state fips codes
state_fips <- fips_codes %>% select(state_name, state_code) %>% 
  distinct() %>% rename(state = state_name)

rm(aeppPCs)
```

## Built capital

Built capital includes: foodbev_est_CBP, est_CBP, broad, broad_11, highway_km, pc1b_manufacturing, and pc2b_infrastructure.

broad, broad_11, and highway_km are not able to be presented for state/national comparisons. Both broadband measures are already per capita and highway_km is calculated such that aggregating up does not make sense. 

```{r built}

# Select data of interest for county level
built_county <- capitals %>% select(
  fips, foodbev_est_CBP_15, est_CBP_15, 
  foodbev_est_CBP_10, est_CBP_10,
  broad, broad_prct, broad_11, highway_km, highway_popwtdist,
  pop_15_CBP, pop_10_CBP_new, 
  pc1b_manufacturing, pc2b_infrastructure)

# Summarise data at the state level
built_state <- capitals %>% group_by(state) %>% select(
  foodbev_est_CBP_15, est_CBP_15, foodbev_est_CBP_10, est_CBP_10,
  pop_15_CBP, pop_10_CBP_new) %>% summarise_all(
    ~sum(.x, na.rm = TRUE)) %>% mutate(across(
      state, ~str_to_title(.x))) %>% left_join(state_fips) %>% select(
        state_code, foodbev_est_CBP_15:pop_10_CBP_new)  %>% rename(
          fips = state_code)

# Summarise data at the US level
built_us <- built_state %>% summarise(across(!fips, ~sum(.x))) %>% mutate(
  fips = "00")

 # Bind data so we have one data frame with US, state, and county level data
built<- bind_rows(built_county, built_state, built_us)
rm(built_county, built_state, built_us)

# Express all variables in per-capita or per-square mile (code taken from AEPP do file)
built <- built %>% mutate(
  foodbev_est_CBP_15=10000*foodbev_est_CBP_15/pop_15_CBP,
  est_CBP_15=10000*est_CBP_15/pop_15_CBP, 
  foodbev_est_CBP_10=10000*foodbev_est_CBP_10/pop_10_CBP_new,
  est_CBP_10=10000*est_CBP_10/pop_10_CBP_new,
  broad=100*broad_prct,
  broad_11 = 100*broad_11,
  highway_km=1/(highway_popwtdist/1000)) %>% rename(
    broad_16 = broad)

# Add year variable
built1 <- built %>% select(fips, foodbev_est_CBP_10, foodbev_est_CBP_15) %>% pivot_longer(
  cols = !fips, 
  names_to = "year", 
  names_prefix = "foodbev_est_CBP_",
  values_to = "foodbev_est_CBP") 

built2 <- built %>% select(fips, est_CBP_10, est_CBP_15) %>% pivot_longer(
  cols = !fips, 
  names_to = "year", 
  names_prefix = "est_CBP_", 
  values_to = "est_CBP")

built3 <- built %>% select(fips, broad_16) %>% mutate(
  year = "16")

built4 <- built %>% select(fips, broad_11) %>% mutate(
  year = "11")

built5 <- built %>% select(fips, highway_km) %>% mutate(
  year = "07")

built6 <- built %>% select(fips, pc1b_manufacturing, pc2b_infrastructure) %>% mutate(
  year = NA)

# Join data
built <- purrr::reduce(list(built1, built2, built3, built4, built5, built6), dplyr::full_join, by = c("fips", "year")) %>% arrange(year, fips) 

built <- built %>% mutate(
  year = str_pad(year, side = "left", pad = "0" , width = 3), 
  year = str_pad(year, side = "left", pad = "2", width = 4))

rm(built1, built2, built3, built4, built5, built6)

# Pivot longer
built <- built %>% pivot_longer(
  cols = !c(fips, year), 
  values_to = "value", 
  names_to = "variable_name") %>% filter(!is.na(value))

# Create column topic area 
built <- built %>% mutate(
  topic_area = ifelse(
    variable_name == "est_CBP" | 
      variable_name =="foodbev_est_CBP", "Food Processors", "Infrastructure"))

```

## Cultural capital

Cultural capital includes create_jobs, racial_div, pub_lib, create_indus_09, create_indus_14, museums, pc1c_artsdiversity and pc2c_creativeindustries. 

The following are NOT calculated at the state/national level. 
racial_div is not seeming to work well at the state/national level. The variable is divided by 833.33 and maybe this is what is making it not work. pub_lib, create_indus_09, create_indus_14, and museums cannot be aggregated up to the state/national level because they are per 10,000 people. 

```{r cultural}

# Select variables to be used in county level 
cultural_county <- capitals %>% select(
  fips, create_jobs, total_emp, White_not_Latino2010,African_American2010,
  Native_American2010, Asian_American2010, Other2010, Latino2010, pub_lib, create_indus_09, create_indus_14, museums, pc1c_artsdiversity, pc2c_creativeindustries)

# Create state-level variables
cultural_state <- capitals %>% group_by(state) %>% select(
  state, create_jobs, total_emp) %>% summarise_all(
    ~sum(.x, na.rm = TRUE)) %>% mutate(across(
      state, ~str_to_title(.x))) %>% left_join(state_fips) %>% select(
        state_code, create_jobs, total_emp)  %>% rename(
          fips = state_code) 

# Create national-level variables
cultural_us <- cultural_state %>% summarise(across(!fips, ~sum(.x))) %>% mutate(
  fips = "00")

 # Bind data so we have one data frame with US, state, and county level data
cultural <- bind_rows(cultural_county, cultural_state, cultural_us)
rm(cultural_county, cultural_state, cultural_us)

# Express variables in per-capita or per-square mile (code taken from AEPP do file)
cultural <- cultural  %>%  mutate(
  create_jobs=100*create_jobs/total_emp,
  racial_div = (10000- (White_not_Latino2010^2 +	African_American2010^2 + Native_American2010^2 + Asian_American2010^2 + Other2010^2 + Latino2010^2))/833.333333333333)

cult1 <- cultural %>% select(fips, create_jobs) %>% mutate(
  year = "2013")
cult2 <- cultural %>% select(fips, racial_div) %>% mutate(
  year = "2010")
cult3 <- cultural %>% select(fips, pub_lib) %>% mutate(
  year = "2012")
cult4 <- cultural %>% select(fips, create_indus_09, create_indus_14) %>% pivot_longer(
  cols = !fips, 
  names_to = "year", 
  names_prefix = "create_indus_",
  values_to = "create_indus") %>% mutate(
    year = str_pad(year, side = "left", pad = "0" , width = 2),
    year = str_pad(year, side = "left", pad = "0" , width = 3),
    year = str_pad(year, side = "left", pad = "2", width = 4))

cult5 <- cultural %>% select(fips, museums) %>% mutate(
  year = "2015")
cult6 <- cultural %>% select(fips, pc1c_artsdiversity, pc2c_creativeindustries) %>% mutate(
  year = NA)

# Join data
cultural <- purrr::reduce(list(cult1,cult2, cult3, cult4, cult5, cult6), dplyr::full_join, by = c("fips", "year")) %>% arrange(year, fips) %>% select(fips, year, everything())

rm(cult1,cult2, cult3, cult4, cult5, cult6)

# Pivot longer
cultural <- cultural %>% pivot_longer(
  cols = !c(fips, year), 
  values_to = "value", 
  names_to = "variable_name") %>% filter(!is.na(value))

# Add topic area
cultural <- cultural %>% mutate(
  topic_area = ifelse(
    variable_name=="racial_div", "Population", "Community Resources"))

```

## Financial capital 

Financial capital includes localgovfin_07, localgovfin_12, 
owner_occupied_2010, owner_occupied_2012, and pc1f. 

```{r financial}

# Financial variables at the county level
financial_county <- capitals %>% select(
  fips, localgovfin_07, localgovfin_12, 
  deposits, owner_occupied_2010, owner_occupied_2012,
  pop_10_CBP_new, pop_12, pop_15_CBP, pc1f)

# Financial variables at the state level
financial_state <- capitals %>% group_by(state) %>% select(
  localgovfin_07, localgovfin_12, 
  deposits, owner_occupied_2010, owner_occupied_2012,
  pop_10_CBP_new, pop_12, pop_15_CBP) %>% summarise_all(
    ~sum(.x, na.rm = TRUE)) %>% mutate(across(
      state, ~str_to_title(.x))) %>% left_join(state_fips) %>% select(
        state_code, localgovfin_07:pop_15_CBP)  %>% rename(
          fips = state_code)

# Summarise data at the US level
financial_us <- financial_state %>% summarise(across(!fips, ~sum(.x))) %>% mutate(
  fips = "00")

 # Bind data so we have one data frame with US, state, and county level data
financial <- bind_rows(financial_county, financial_state, financial_us)
rm(financial_county, financial_state, financial_us)


# Express variables in per-capita or per-square mile (code taken from AEPP do file)
financial <- financial %>% mutate(
  localgovfin_07 = localgovfin_07/pop_10_CBP_new, 
  localgovfin_12 = localgovfin_12/pop_12,
  deposits = deposits/pop_15_CBP,
  owner_occupied_2010 = owner_occupied_2010/pop_10_CBP_new, 
  owner_occupied_2012 = owner_occupied_2012/pop_12)

# Add year variables
fin1 <- financial %>% select(
  fips, localgovfin_07, localgovfin_12) %>% pivot_longer(
    cols = !fips, 
    names_to = "year", 
    names_prefix = "localgovfin_",
    values_to = "localgovfin")  %>% mutate(
  year = str_pad(year, side = "left", pad = "0" , width = 3), 
  year = str_pad(year, side = "left", pad = "2", width = 4))

fin2 <- financial %>% select(
  fips, owner_occupied_2010, owner_occupied_2012) %>% pivot_longer(
    cols = !fips, 
    names_to = "year", 
    values_to = "owner_occupied", 
    names_prefix = "owner_occupied_")

fin3 <- financial %>% select(fips, deposits) %>% mutate(
  year = "2016")

fin4 <- financial %>% select(fips, pc1f) %>% mutate(
  year = NA)

# Join data
financial <- purrr::reduce(list(fin1, fin2, fin3, fin4), dplyr::full_join, by = c("fips", "year")) %>% arrange(year, fips) 
rm(fin1, fin2, fin3, fin4)

# Pivot longer
financial <- financial %>% pivot_longer(
  cols = !c(fips, year), 
  values_to = "value", 
  names_to = "variable_name") %>% filter(!is.na(value)) %>% mutate(
    topic_area = "Financial")

```

## Human capital 

Human capital includes ed_attain_10, ed_attain_15, health_factors, health_outcomes, rood_secure_rev, food_secure_10, insured_13, insured_10,  primary_care_17,  primary_care_10, pc1h_healtheducation, pc2h_medicalfoodsecurity. Note, food_secure_rev is from 2014

Note that ed_attain for 2010 is already a percentage, so don't need to divide by population. 

Variables we cannot provide state and national totals include ed_attain_10, food_secure_rev, food_secure_10, insured_13, insured_10 as they are provided as percentages. 

```{r human}

# County level data 
human_county <- capitals %>% select(
  fips,
  ed_attain_10, ed_attain_15, 
  food_secure_rev, food_secure_10, 
  insured_13, insured_10,
  primary_care_17, primary_care_10,
   pop_10_CBP_new, pop_15_CBP, adult_pop_15, 
  pc1h_healtheducation, pc2h_medicalfoodsecurity)

# State level data
human_state <- capitals %>% group_by(state) %>% select(
  ed_attain_15, primary_care_17, primary_care_10) %>% summarise_all(
    ~sum(.x, na.rm = TRUE)) %>% mutate(across(
      state, ~str_to_title(.x))) %>% left_join(state_fips) %>% select(
        state_code, ed_attain_15, primary_care_17, primary_care_10)  %>% rename(
          fips = state_code)

# Summarise data at the US level
human_us <- human_state %>% summarise(across(!fips, ~sum(.x))) %>% mutate(
  fips = "00")

 # Bind data so we have one data frame with US, state, and county level data
human <- bind_rows(human_county, human_state, human_us)
rm(human_county, human_state, human_us)

# Express variables in per-capita or per-square mile (code taken from AEPP do file)
human <- human %>% mutate(
  ed_attain_10 = ed_attain_10, 
  ed_attain_15 = 100*ed_attain_15/adult_pop_15, 
  food_secure_rev=100*food_secure_rev, 
  food_secure_10 = 100*food_secure_10, 
  insured_13 = 100*insured_13, 
  insured_10 = 100*insured_10, 
  primary_care_17=10000*primary_care_17/pop_15_CBP, 
  primary_care_10=10000*primary_care_10/pop_10_CBP_new) %>% 
  rename(food_secure_14 = food_secure_rev)

hum1 <- human %>% select(fips, ed_attain_10, ed_attain_15) %>% pivot_longer(
  cols = !fips, 
  names_to = "year", 
  names_prefix = "ed_attain_", 
  values_to = "ed_attain")

hum2 <- human %>% select(fips, food_secure_10, food_secure_14) %>% pivot_longer(
  cols = !fips, 
  names_to = "year", 
  names_prefix = "food_secure_", 
  values_to = "food_secure")

hum3 <- human %>% select(fips, insured_10, insured_13) %>% pivot_longer(
  cols = !fips, 
  names_to = "year", 
  names_prefix = "insured_", 
  values_to = "insured")

hum4 <- human %>% select(fips, primary_care_10, primary_care_17) %>% pivot_longer(
  cols = !fips, 
  names_to = "year", 
  names_prefix = "primary_care_", 
  values_to = "primary_care")

hum5 <- human %>% select(fips, pc1h_healtheducation, pc2h_medicalfoodsecurity) %>% mutate(
  year = NA)

# Join data
human <- purrr::reduce(list(hum1, hum2, hum3, hum4, hum5), dplyr::full_join, by = c("fips", "year")) %>% arrange(year, fips) 

human <- human %>% mutate(
  year = str_pad(year, side = "left", pad = "0" , width = 3), 
  year = str_pad(year, side = "left", pad = "2", width = 4))

#write_csv(human, "community_wealth/human_capital.csv")
rm(hum1, hum2, hum3, hum4, hum5)

# Pivot longer
human <- human %>% pivot_longer(
  cols = !c(fips, year), 
  values_to = "value", 
  names_to = "variable_name") %>% filter(!is.na(value)) 

# Add topic areas
human <- human %>% mutate(
  topic_area = case_when(
    variable_name=="pc1h_healtheducation" |
      variable_name=="pc2h_medicalfoodsecurity" |
    variable_name=="ed_attain" |
      variable_name=="insured" ~ "Population", 
    variable_name=="food_secure" ~ "Food Insecurity", 
    variable_name=="primary_care" ~ "Community Resources"))

```

## Natural capital 

Natural capital includes natamen_scale, prime_farmland, conserve_acre, acre_FSA, acre_NFS, pc1n_naturalamenitiesconservation, and pc2n_farmland. 


natamen_scale not provided at the state/national level. 

```{r natural}

# County level data
natural_county <- capitals %>% select(
  fips, natamen_scale, prime_farmland, conserve_acre, acre_FSA, acre_NFS, pc1n_naturalamenitiesconservation, pc2n_farmland, 
  acres, acre_all)

# State level data
natural_state <- capitals %>% group_by(state) %>% select(
  prime_farmland, conserve_acre, acre_FSA, acre_NFS,  
  acres, acre_all) %>% summarise_all(
    ~sum(.x, na.rm = TRUE)) %>% mutate(across(
      state, ~str_to_title(.x))) %>% left_join(state_fips) %>% select(
        state_code, prime_farmland:acre_all)  %>% rename(
          fips = state_code)

# Summarise data at the US level
natural_us <- natural_state %>% summarise(across(!fips, ~sum(.x))) %>% mutate(
  fips = "00")

 # Bind data so we have one data frame with US, state, and county level data
natural <- bind_rows(natural_county, natural_state, natural_us)
rm(natural_county, natural_state, natural_us)

# Express variables in per-acre (code taken from AEPP do file)
natural <- natural %>% mutate(
  prime_farmland=100*prime_farmland/acres, 
  conserve_acre=100*conserve_acre/acre_all, 
  acre_FSA = 100*acre_FSA/acre_all, 
  acre_NFS = 100*acre_NFS/acre_all)

nat1 <- natural %>% select(fips, prime_farmland) %>% mutate(year="2012")
nat2 <- natural %>% select(fips, conserve_acre) %>% mutate(year="2016")
nat3 <- natural %>% select(fips, acre_FSA, acre_NFS) %>% mutate(year = "2017") 
nat4 <- natural %>% select(fips, natamen_scale) %>% mutate(year = "1999")
nat5 <- natural %>% select(fips, pc1n_naturalamenitiesconservation, pc2n_farmland) %>% mutate(year = NA)

# Join data
natural<- purrr::reduce(list(nat1, nat2, nat3, nat4, nat5), dplyr::full_join, by = c("fips", "year")) %>% arrange(year, fips) %>% select(fips, year, everything())

#write_csv(natural, "community_wealth/natural_capital.csv")
rm(nat1, nat2, nat3, nat4, nat5) 

# Pivot longer
natural <- natural %>% pivot_longer(
  cols = !c(fips, year), 
  values_to = "value", 
  names_to = "variable_name") %>% filter(!is.na(value)) 

# Add topic areas
natural <- natural %>% mutate(
  topic_area = ifelse(
    variable_name=="prime_farmland", "Agriculture", "Natural"))
```

## Social-political capital 

Social political capital includes nccs09, nccs14, pvote08, pvote12, respn10, assn09, assn14, 
pc1s_nonprofitsocialindustries, and pc2s_publicvoiceparticipation.  


```{r socpol}

# County level data 
socpol_county <- capitals %>% select(
  fips, nccs09, nccs14, pvote08, pvote12, respn10, assn09, assn14, 
pc1s_nonprofitsocialindustries, pc2s_publicvoiceparticipation, 
pop09,pop_14)

socpol_state <- capitals %>% group_by(state) %>% select(
  nccs09, nccs14, pop09, pop_14)%>% summarise_all(
    ~sum(.x, na.rm = TRUE)) %>% mutate(across(
      state, ~str_to_title(.x))) %>% left_join(state_fips) %>% select(
        state_code, nccs09:pop_14)  %>% rename(
          fips = state_code)

# Summarise data at the US level
socpol_us <- socpol_state %>% summarise(across(!fips, ~sum(.x))) %>% mutate(
  fips = "00")

 # Bind data so we have one data frame with US, state, and county level data
socpol <- bind_rows(socpol_county, socpol_state, socpol_us)
rm(socpol_county, socpol_state, socpol_us)

# Make per capita, etc. (code taken from AEPP do file)
socpol <- socpol %>% mutate(
  nccs09 = 1000*nccs14/pop09, 
  nccs14 = 1000*nccs14/pop_14,
  pvote08 = 100*pvote08, 
  pvote12 = 100*pvote12, 
  respn10 = 100*respn10)

sp1 <- socpol %>% select(fips, nccs09, nccs14) %>% pivot_longer(
  cols = !fips, 
  names_to = "year", 
  names_prefix = "nccs", 
  values_to = "nccs")

sp2 <- socpol %>% select(fips, pvote08, pvote12) %>% pivot_longer(
  cols = !fips, 
  names_to = "year", 
  names_prefix = "pvote", 
  values_to = "pvote")

sp3 <- socpol %>% select(fips, respn10) %>% mutate(
  year = "10") %>% rename(respn = respn10)

sp4 <- socpol %>% select(fips, assn09, assn14) %>% pivot_longer(
  cols = !fips, 
  names_to = "year", 
  names_prefix = "assn", 
  values_to = "assn")

sp5 <- socpol %>% select(fips, pc1s_nonprofitsocialindustries, pc2s_publicvoiceparticipation) %>% mutate(year = NA)
  
# Join data
socpol <- purrr::reduce(list(sp1, sp2, sp3, sp4, sp5), dplyr::full_join, by = c("fips", "year")) %>% arrange(year, fips) 

socpol <- socpol %>% mutate(
  year = str_pad(year, side = "left", pad = "0" , width = 3), 
  year = str_pad(year, side = "left", pad = "2", width = 4))

rm(sp1, sp2, sp3, sp4, sp5)

# Pivot longer
socpol <- socpol %>% pivot_longer(
  cols = !c(fips, year), 
  values_to = "value", 
  names_to = "variable_name") %>% filter(!is.na(value)) 

# Add topic area
socpol <- socpol %>% mutate(
  topic_area = case_when(
    variable_name=="pvote" |
      variable_name=="respn"| 
      variable_name=="pc2s_publicvoiceparticipation" ~ "Population", 
    variable_name=="nccs" |
      variable_name=="assn" | 
      variable_name=="pc1s_nonprofitsocialindustries" ~ "Community Resources"))


```

## Bind all data and add county and state names

```{r}

# Bind data
community_wealth <- bind_rows(built, cultural, financial, human, natural, socpol)

# Add state and county name
fips_codes <- fips_codes %>% select(state_code, county_code, everything()) %>% unite("fips", state_code:county_code, sep = "") %>% select(-state) %>% rename(county_name = county)

community_wealth <- left_join(community_wealth, fips_codes)

# Add category
community_wealth <- community_wealth %>% mutate(
  category = case_when(
    topic_area=="Agriculture" | 
      topic_area=="Community Resources" | 
      topic_area=="Financial" |
      topic_area=="Infrastructure" | 
      topic_area=="Natural" |
      topic_area=="Population" ~ "Community Characteristics", 
   topic_area=="Food Insecurity" ~ "Food Access", 
   topic_area=="Food Processors" ~ "Processing & Distribution"))

# Re-order data
community_wealth <- community_wealth %>% select(
  fips, county_name, state_name, category, topic_area, year, variable_name, value)

write_csv(community_wealth, "community_wealth.csv")
```